---
title: "Deaths related to Cancer in Georgia from 2009 to 2020"
author: "Iris Cho"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: scroll
---

```{r setup, include=FALSE}
library(highcharter)
library(dplyr)
library(tidyverse)
library(here)
library(viridisLite)
library(reticulate)
library(flexdashboard)
library(shiny)
library(plotly)
library(sf)
library(geojsonio)
library(tigris)

# merge and clean data
file_path <- ("~/Documents/Emory RSPH/S2025/Current Topics in DS/WONDER_GA/")
files <- list.files(path = file_path, pattern = "ga.*\\.csv", full.names = TRUE)
dfs <- lapply(files, function(file) {
  df <- read.csv(file)
  required_cols <- c("YEAR", "COUNTYFIPS", "COUNTY", "STATE", "TOT_DEATH")
  missing_cols <- setdiff(required_cols, colnames(df))
  df[missing_cols] <- NA
  df <- df[required_cols]
  return(df)
})

master_df <- bind_rows(dfs)
master_df$TOT_DEATH <- as.numeric(master_df$TOT_DEATH)

path <- ("~/Documents/Emory RSPH/S2025/Current Topics in DS/WONDER_GA/")
ga_files <- list.files(path = path, pattern = "\\.csv$", full.names = TRUE)
combined_data <- map_df(ga_files, read_csv)

combined_data$TOT_DEATH <- as.numeric(combined_data$TOT_DEATH)
ga_map <- counties(state = "GA", year = 2020, class = "sf")

merged_map <- merge(ga_map, combined_data, by.x = "GEOID", by.y = "COUNTYFIPS")
merged_map$GEOID <- as.character(merged_map$GEOID)
```

Visualizations
======================================================================
Row {data-width=700}
-----------------------------------------------------------------------
### Top 10 Counties Death Trend (2009–2020)
##### Key Takeaway: This plot highlights the top 20 counties with the highest cancer-related mortality rates. Notably, counties within metropolitan areas remain in the top 10, which may be influenced by healthcare accessibility and high population density in these counties.
```{r, time-series, echo=FALSE, message=FALSE}

top_10_counties <- master_df |>
  group_by(COUNTY) |>
  summarise(Total_Deaths = sum(TOT_DEATH, na.rm = TRUE)) |>
  arrange(desc(Total_Deaths)) |>
  slice(1:10) |>
  pull(COUNTY)

death_trend <- master_df |>
  group_by(YEAR, COUNTY) |>
  summarise(Total_Deaths = sum(TOT_DEATH, na.rm = TRUE), .groups = "drop") |>
  mutate(Visible = ifelse(COUNTY %in% top_10_counties, TRUE, FALSE))

p <- plot_ly()

for (County in unique(death_trend$COUNTY)) {
  p <- p |>
    add_trace(
      data = subset(death_trend, COUNTY == County),
      x = ~YEAR,
      y = ~Total_Deaths,
      type = 'scatter',
      mode = 'lines+markers',
      name = County,
      visible = ifelse(County %in% top_10_counties, TRUE, "legendonly")  # Top 10 visible, others hidden
    )
}

p <- p |>
  layout(
    title = "Cancer-Related Deaths by County 2009 to 2020",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Total Deaths"),
    hovermode = "x")
p

```

Row {data-width=700}
----------------------------------------------------------------------------
### Cancer-Related Deaths by County in Georgia (2020)
##### Key Takeaway: This spatial heatmap visually represents the spatial distribution of cancer-related deaths in 2020, helping viewers identify regions with high population concentrations. Counties without color indicate suppressed or underreported data which indicates supplemental intervention is needed. 
```{r}
geojson_file <- tempfile(fileext = ".geojson")
st_write(merged_map, geojson_file, driver = "GeoJSON")
ga_geojson <- geojson_read(geojson_file, what = "sp")
ga_geojson_list <- geojson_list(ga_geojson)

colstops <- list(
  list(0, "#440154"),  
  list(0.25, "#3b528b"),
  list(0.5, "#21918c"),  
  list(0.75, "#5ec962"),  
  list(1, "#fde725")
)

highchart(type = "map") |>
  hc_add_series_map(
    map = ga_geojson_list,  
    df = merged_map,  
    name = "Deaths",   
    value = "TOT_DEATH",
    joinBy = list("GEOID", "GEOID"), 
    dataLabels = list(enabled = TRUE, format = '{point.properties.NAME}')) |>
  hc_colorAxis(stops = colstops) |>
  hc_legend(valueDecimals = 0, valueSuffix = " deaths") |>
  hc_mapNavigation(enabled = TRUE) |>
  hc_add_theme(hc_theme_flat()) |>
  hc_title(
    text = "Cancer-Related Deaths by County in Georgia",
    style = list(fontSize = "18px", fontWeight = "bold", color = "#333333")
  )
```

Row
-----------------------------------------------------------------------
#### Purpose of the Dashboard
##### The purpose of this dashboard is to provide a dynamic, visual exploration of cancer-related mortality trends across Georgia counties from 2009 to 2020. It aims to highlight regional disparities in cancer burden, assist in public health decision-making, and offer a user-friendly tool for academic, clinical, and policy audiences. 

About
============================================================================
This project uses the AHRQ (Agency for Healthcare Research and Quality)’s data for Social Determinants of Health (SDOH) and CDC’s WONDER (Wide-ranging Online Data for Epidemiologic Research) to understand cancer mortality rates and it’s relation to SDOH at county-level in Georgia from 2009 to 2020. The data were aggregated from sources like the U.S. Census Bureau, CDC, and Healthcare Cost and Utilization Project (HCUP) using surveys, administrative records, disease surveillance and death certificates. The study population includes residents of Georgia in 159 counties from 2009 to 2020 who’s cause of death was determined to be cancer. Key variables include total deaths in the county related to cancer, counties, year reported, total population of the county, median age, race, household factors, and healthcare access.
AHRQ Database: https://www.ahrq.gov/sdoh/data-analytics/sdoh-data.html
CDC WONDER: https://wonder.cdc.gov/cancer.html



